The brightness and spatial distributions of terrestrial radio sources

### 2013-01-08 articles/Article-RFI-distribution/article-rfi-distribution.tex

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
\end{abstract}

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters. For deliberate transmitters, the frequency at which interference occurs can identify the involved class of devices, because devices are constrained by the bands that have been allocated for the given class.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

We assume that the source observed is fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to predict the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - c\pi r^2$, with $N$ the total number of sources.

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{c\pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{c\pi Ig}{S^2}.
\end{align}

@@

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a hemisphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.

@@

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $c$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

%\begin{figure}[tb]
%\begin{center}
%\includegraphics[width=12cm]{img/plot-passband.pdf}
%\caption{Effect of the low and high-band frequency responses on the amplitude histogram for different distributions.}
%\label{fig:plot-passband-histogram}
%\end{center}
%\end{figure}

@@

%In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given. To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the band-pass. The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference. As expected, it can still be determined that the intrinsic distribution is a -2 power law.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = \zeta Ig / r^\alpha$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment} % TODO correct citation if full citation available
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. % TODO
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

To detect RFI, the AOFlagger \citep{post-correlation-rfi-classification, LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used. However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector. Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions. This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.

>>>>

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent LOFAR from detecting the EoR.
\end{abstract}

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kinds of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters. For deliberate transmitters, the frequency at which interference occurs can identify the involved class of devices, because devices are constrained by the bands that have been allocated for the given class.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

We assume that the source observed is fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50 km.

Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
\end{equation}
for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to predict the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{\rho \pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}(r)$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho \pi Ig}{S^2}.
\end{align}

@@

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
% (c=rho)
\begin{align}
f_p(S)
= & \frac{d}{dS} \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.

@@

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

@@

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = \zeta Ig / r^\alpha$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}.
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used. However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector. Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions. The variation could be caused by the different power law exponents that source populations at different frequencies might have. It could also be caused by approximately similar underlying power-law distribution that only differ in the number of transmitters. In that case, the underlying power law might not yet be apparent, because not enough samples are combined.
